The zero--one laws of Kolmogorov and Hewitt--Savage in categorical probability

### 2020-05-03 categorical-zero-one.tex
acknowledged Kenta
Some rev1 edits, added list file

\paragraph*{Acknowledgments.} We thank Paolo Perrone for discussions and \Cref{poset_case}.

@@

Returning to the general theory, in terms of string diagrams the unique morphism $\discard{X} : X \to I$ for an object $X \in \cC$ is denoted by
\[
	\tikzfig{terminal}
\]
The projection maps $\id \otimes \discard{Y} : X \otimes Y \to X$ and $\discard{Y} \otimes \id : X \otimes Y \to Y$, which are the ones from~\eqref{projections}, are correspondingly written as
\[
	\tikzfig{marginals}
\]

@@

It is known that $\Stoch$ is causal~\cite[Example~11.34]{markov_cats}, and therefore so are the subcategories $\BorelStoch$ and $\FinStoch$. We do not know whether $\CRing_+\op$ is causal, although we suspect that~\cite[Example~11.31]{markov_cats} can be modified to show that it is not.

@@

\begin{definition}
	\label{defnasequal}
	Given morphisms $p : A \to X$ and $f,g : X \to Y$ in a Markov category, we say that $f$ and $g$ are \emph{$p$-a.s.~equal} if
	\[
		\tikzfig{asequal}
	\]
	and we also write this more concisely as $f =_{p\as} g$.
\end{definition}

@@

Using the multiplicativity of the comonoid structure of~\eqref{multiplicativity}, we see that \Cref{two_infproducts,infprod_operadic} hold also for Kolmogorov products.

@@

\begin{example}
    \label{setmulti}
    Let $\SetMulti$ be the category with objects ordinary sets, morphisms $f: X \to Y$ given by \emph{multivalued functions}, which are maps $f: X \to \cP(Y)\setminus \emptyset$, with composition of morphisms $f : X \to Y$ and $g : Y \to Z$ given by
    	\begin{equation}
		\label{relational_composition}
		(g\circ f)(x) := \bigcup_{y \in f(x)} g(y).
	\end{equation}
    See also \cite[Example~2.6]{markov_cats} for a treatment of the full subcategory $\FinSetMulti$ on finite sets.
    The Cartesian product of sets gives it a symmetric monoidal structure, if for $f : X_1 \to Y_1$ and $g: X_2 \to Y_2$ we set
	\[
	(f \tensor g)(x_1,x_2) := f(x_1) \times g(x_2) \:\subseteq\: Y_1 \times Y_2.
	\]
	With the obvious structure isomorphisms inherited from $\Set$, this turns $\SetMulti$ into a semicartesian symmetric monoidal category, and we can give it a Markov category structure by setting
    $\cop{X}(x) := \{(x,x)\} \subset X \times X$.
    Then the deterministic morphisms are precisely those sending each point into a singleton set, resulting in an isomorphism of categories
    $\SetMulti_\detc \cong \sf{Set}$.

@@

\begin{lemma}[The infinite independence lemma]
	Suppose $p: A \to X_J$ displays the conditional independence $\perp_i X_i \mid\mid A$.
	Then for every $i \in J$, it also displays the independence $X_i \perp X_{J \setminus \{i\}} \mid\mid A$.
\end{lemma}
\begin{proof}
	This is nontrivial because the assumption by definition only involves conditional independence of the finite marginalizations. What we need to prove is that
	\[
		\tikzfig{infind_lemma_goal}
	\]
	As the codomain is exactly the Kolmogorov product $X_J$, it is enough to prove this equation upon composing the second output with the finite marginalizations $\pi_F : X_{J\setminus\{i\}} \to X_F$. But this is a consequence of the assumption.
\end{proof}

@@

\begin{theorem}[Abstract Kolmogorov zero--one law]
    \label{thm:kolmog}
    Suppose that $\cC$ is a Markov category, and that $(X_i)_{i \in J}$ is a family of objects with Kolmogorov product $X_J$ and finite marginalizations $\pi_F : X_J \to X_F$. Suppose that morphisms $p: A \to X_J$ and deterministic $s : X_J \to T$ satisfy the following:
    \begin{enumerate}
	    \item $p$ displays the conditional independence $\perp_i X_i \mid\mid A$.
	    \item For every finite $F \subseteq J$, the joint
    		\[
	    		\tikzfig{kolmog_joint}
    		\]
    		displays the conditional independence $X_F \perp T \mid\mid A$.
    \end{enumerate}
Then the composite $sp : A \to T$ is deterministic as well.
\end{theorem}
Note that this result applies even in the case where $J$ is finite, in which case it amounts to \Cref{determinismlemma}.
\begin{proof}
	Assuming $\ast \not \in J$, let us write $X_\ast := T$, so that $X_J \otimes X_\ast$ is the Kolmogorov product of the family $(X_i)_{i \in J \,\sqcup\,\{\ast\}}$. It is clear from the definitions and the assumption that the morphism
    \[
	\tikzfig{kolmog_joint2}	    
    \]
    displays the conditional independence $\perp_{i \in J\,\sqcup\,\{\ast\}} X_i \mid\mid A$.
    By the infinite independence lemma, this means that it also displays the conditional independence $X_\ast \perp X_J \mid\mid A$.
    Now the determinism lemma implies exactly what we want, namely that the composite $sp : A \to T$ is deterministic.
\end{proof}

@@

@@

\begin{corollary}[Hewitt-savage zero--one law]
    Suppose $\Omega$ is a standard Borel space with probability measure $P$, and let $\left(f_i: \Omega \to X\right)_{i\in \bN}$ be a sequence of independent, identically distributed random variables taking values in a standard Borel space $X$.
    Suppose $S \subseteq \prod_{i \in \bN} X = X_\bN$ is measurable in the product $\sigma$-algebra, and suppose moreover that, for each finite permutation $\sigma: \bN \to \bN$, we have $\hat{\sigma}(S) = S$ (in other words, $S$ is invariant under finite permutations of the variables).
    Then $P(f_\bN\inv(S)) \in \{0,1\}$, where $f_\bN$ again denotes the induced map $\Omega \to X_\bN$.
\end{corollary}
\begin{proof}
    The proof is similar to the previous proof. This time we want to apply \Cref{thm:hewsav} with $\cC = \BorelStoch$, $J = \bN$, $A = I$.
    Again, we have $P: I \to \Omega$.
    As before, we let $s: X_\bN \to \{0,1\}$ be the indicator function of $S$, and we let $p = f_\bN P$.
    The i.i.d.~assumption on the $f_j$ implies that $\hat{\sigma}p = p$, and the assumption on $S$ implies $s\hat{\sigma} = s$ for finite permutations $\sigma$.
    Hence we can apply \Cref{thm:hewsav}. This again shows that the map $sp: I \to \{0,1\}$ is deterministic, which, just as before, means $P(f_\bN\inv(S)) \in \{0,1\}$.
\end{proof}

@@

\begin{example}
	Let $A = \{*\}$ and $X = Y = Z = [0,1]$ with the usual topology.
    Let $f: A \to HX$ send $*$ to $X$, and define
    \[g(x) :=  \begin{cases} [0,1/2] & x < \frac{1}{3}\\ \{1/2\} & \frac{1}{3} < x < \frac{2}{3}\\ [1/2,1] & x > \frac{1}{2}\end{cases}\]
    \[
    	h_1(y) := [0,1], \qquad h_2(y) = \begin{cases}[0,1] & y \neq \frac{1}{2}\\ \{0\} &  y = \frac{1}{2}\end{cases}
    \]
    Using the definition of the hyperspace topology, it is straightforward to see that these maps are continuous. To check the hypothesis of the statement of causality, we must verify that the resulting two maps $A \to Y \times Z$ agree; these are just closed subsets of $[0,1]^2$. From the definition of the composition, we see that any point $(y,z)$ with $y \in g(x)$ for some $x$ and $z \in h_1(y)$ must lie in the set corresponding to $h_1$,
and similarly for $h_2$.
Hence the set corresponding to $h_1$ contains all points, while the set corresponding to $h_2$ contains all points with $y \neq 1/2$.
But the closure of $\{(y,z) \mid y \neq 1/2\}$ in $[0,1]^2$ is the whole space, so that both sets are indeed equal to the whole space.

@@

Strangely enough, the only examples that we know of so far are of two kinds: first, cartesian monoidal categories (meaning that $\cC = \cC_\det$) with all products; and second, the finitary complete monoids of \Cref{poset_case} which are monoidal posets. For the former, the universal property of the monoidal structure implies that the causality axiom holds. For the latter, it holds trivially since any two parallel morphisms are equal.

>>>>

\paragraph*{Acknowledgments.} We thank Paolo Perrone for discussions and \Cref{poset_case} as well as Kenta Cho for comments on an earlier version of this paper.

@@

Returning to the general theory, in terms of string diagrams the unique morphism $\discard{X} : X \to I$ for an object $X \in \cC$ is denoted by
\[
	\tikzfig{terminal}
\]
The projection maps $\id \otimes \discard{Y} : X \otimes Y \to X$ and $\discard{X} \otimes \id : X \otimes Y \to Y$, which are the ones from~\eqref{projections}, are correspondingly written as
\[
	\tikzfig{marginals}
\]

@@

It is known that $\Stoch$ is causal~\cite[Example~11.34]{markov_cats}, and
therefore so are the subcategories $\BorelStoch$ and $\FinStoch$.
\todo[inline]{Added as per reviewer's suggestion, but unsure if it's a good
  inclusion or not.}
\begin{example}
$\CRing_{+}^{\op}$ is \emph{not} causal.
To see this, consider the ring $\mathbb{Z}[\epsilon]/(\epsilon^{2})$.
Elements have the form $a + b\epsilon$, with $a,b\in\bZ$, and multiplication is
defined by the convention $\epsilon^{2} = 0$.
Consider the additive maps given by
\begin{equation}
  f(a + b\epsilon) = & b\\
  g(a + b \epsilon) = & b\\
  h_{1}(a+ b \epsilon) = & a + b \epsilon\\
  h_{2}(a + b \epsilon) = & 0
\end{equation}
Since $gf = 0$, it's clear that the hypothesis holds, so if $\CRing_{+}^{\op}$
were causal, we would have the equation
$g(f(h_{1}(x)y)z) = g(f(h_{2}(x)y)z)$ for all
$x,y,z \in \bZ[\epsilon]/(\epsilon^{2})$.
However, this clearly fails if $z = \epsilon$, $y=1$ and $x = \epsilon$.
\end{example}

@@

\begin{definition}
	\label{defnasequal}
	Given morphisms $p : A \to X$ and $f,g : X \to Y$ in a Markov category, we
    say that $f$ and $g$ are \emph{$p$-a.s.~equal} (or $p$-\emph{almost surely}.~equal) if
	\[
		\tikzfig{asequal}
	\]
	and we also write this more concisely as $f =_{p\as} g$.
  \end{definition}

  This term comes from measure theory, and should be interpreted as meaning
  ``for all $a\in A$, if $x \in X$ is sampled according to $p(a)$, the
  probability that $f(x) \neq g(x)$ is zero''.

@@

Using the multiplicativity of the comonoid structure
of~\eqref{multiplicativity}, we see that \Cref{two_infproducts,infprod_operadic}
hold also for Kolmogorov products, in the sense that tensoring together
Kolmogorov products, or taking Kolmogorov products of Kolmogorov products, results
again in a Kolmogorov product.

@@

\begin{example}
    \label{setmulti}
    Let $\SetMulti$ be the category with objects ordinary sets, morphisms $f: X \to Y$ given by \emph{multivalued functions}, which are maps $f: X \to \cP(Y)\setminus \emptyset$, with composition of morphisms $f : X \to Y$ and $g : Y \to Z$ given by
    	\begin{equation}
		\label{relational_composition}
		(g\circ f)(x) := \bigcup_{y \in f(x)} g(y).
      \end{equation}
    $\SetMulti$ could also be described as the Kleisli category of the
    \emph{nonempty powerset monad}, which sends a set $X$ to $\cP(X) \setminus \emptyset$.

    See also \cite[Example~2.6]{markov_cats} for a treatment of the full subcategory $\FinSetMulti$ on finite sets.
    The Cartesian product of sets gives it a symmetric monoidal structure, if for $f : X_1 \to Y_1$ and $g: X_2 \to Y_2$ we set
	\[
	(f \tensor g)(x_1,x_2) := f(x_1) \times g(x_2) \:\subseteq\: Y_1 \times Y_2.
	\]
	With the obvious structure isomorphisms inherited from $\Set$, this turns $\SetMulti$ into a semicartesian symmetric monoidal category, and we can give it a Markov category structure by setting
    $\cop{X}(x) := \{(x,x)\} \subset X \times X$.
    Then the deterministic morphisms are precisely those sending each point into a singleton set, resulting in an isomorphism of categories
    $\SetMulti_\detc \cong \sf{Set}$.

@@

\begin{lemma}[The infinite independence lemma]
  \label{infindeplemma}
	Suppose $p: A \to X_J$ displays the conditional independence $\perp_i X_i \mid\mid A$.
	Then for every $i \in J$, it also displays the independence $X_i \perp X_{J \setminus \{i\}} \mid\mid A$.
\end{lemma}
\begin{proof}
	This is nontrivial because the assumption by definition only involves conditional independence of the finite marginalizations. What we need to prove is that
	\[
		\tikzfig{infind_lemma_goal}
	\]
	As the codomain is exactly the Kolmogorov product $X_J$, it is enough to prove this equation upon composing the second output with the finite marginalizations $\pi_F : X_{J\setminus\{i\}} \to X_F$. But this is a consequence of the assumption.
\end{proof}

@@

\begin{theorem}[Abstract Kolmogorov zero--one law]
    \label{thm:kolmog}
    Suppose that $\cC$ is a Markov category, and that $(X_i)_{i \in J}$ is a family of objects with Kolmogorov product $X_J$ and finite marginalizations $\pi_F : X_J \to X_F$. Suppose that morphisms $p: A \to X_J$ and deterministic $s : X_J \to T$ satisfy the following:
    \begin{enumerate}
	    \item $p$ displays the conditional independence $\perp_i X_i \mid\mid A$.
	    \item For every finite $F \subseteq J$, the joint
    		\[
	    		\tikzfig{kolmog_joint}
    		\]
    		displays the conditional independence $X_F \perp T \mid\mid A$.
    \end{enumerate}
Then the composite $sp : A \to T$ is deterministic as well.
\end{theorem}
Note that this result applies even in the case where $J$ is finite, in which case it amounts to \Cref{determinismlemma}.
\begin{proof}
	Assuming $\ast \not \in J$, let us write $X_\ast := T$, so that $X_J \otimes X_\ast$ is the Kolmogorov product of the family $(X_i)_{i \in J \,\sqcup\,\{\ast\}}$. It is clear from the definitions and the assumption that the morphism
    \[
	\tikzfig{kolmog_joint2}	    
    \]
    displays the conditional independence $\perp_{i \in J\,\sqcup\,\{\ast\}} X_i \mid\mid A$.
    By the infinite independence lemma, \cref{infindeplemma}, this means that it also displays the conditional independence $X_\ast \perp X_J \mid\mid A$.
    Now the determinism lemma implies exactly what we want, namely that the composite $sp : A \to T$ is deterministic.
\end{proof}

@@

We note that the standard Kolmogorov zero--one law holds for any measurable
spaces, not just for standard Borel spaces - hence this is a strictly less
general statement. As discussed in \cref{infprods_stoch}, it seems unlikely that
the general statement can be proven by applying our machinery to the category
$\Stoch$ of measurable spaces and Markov kernels. This leads to the following problem:

@@

\begin{corollary}[Hewitt-savage zero--one law]
    Suppose $\Omega$ is a standard Borel space with probability measure $P$, and let $\left(f_i: \Omega \to X\right)_{i\in \bN}$ be a sequence of independent, identically distributed random variables taking values in a standard Borel space $X$.
    Suppose $S \subseteq \prod_{i \in \bN} X = X_\bN$ is measurable in the product $\sigma$-algebra, and suppose moreover that, for each finite permutation $\sigma: \bN \to \bN$, we have $\hat{\sigma}(S) = S$ (in other words, $S$ is invariant under finite permutations of the variables).
    Then $P(f_\bN\inv(S)) \in \{0,1\}$, where $f_\bN$ again denotes the induced map $\Omega \to X_\bN$.
  \end{corollary}

\begin{proof}
    The proof is similar to the previous proof. This time we want to apply \Cref{thm:hewsav} with $\cC = \BorelStoch$, $J = \bN$, $A = I$.
    Again, we have $P: I \to \Omega$.
    As before, we let $s: X_\bN \to \{0,1\}$ be the indicator function of $S$, and we let $p = f_\bN P$.
    The i.i.d.~assumption on the $f_j$ implies that $\hat{\sigma}p = p$, and the assumption on $S$ implies $s\hat{\sigma} = s$ for finite permutations $\sigma$.
    Hence we can apply \Cref{thm:hewsav}. This again shows that the map $sp: I \to \{0,1\}$ is deterministic, which, just as before, means $P(f_\bN\inv(S)) \in \{0,1\}$.
\end{proof}

@@

\begin{example}
	Let $A = \{*\}$ and $X = Y = Z = [0,1]$ with the usual topology.
    Let $f: A \to HX$ send $*$ to $X$, and define
    \[g(x) :=  \begin{cases} [0,1/2] & x < \frac{1}{3}\\ \{1/2\} & \frac{1}{3} \leq x \leq \frac{2}{3}\\ [1/2,1] & x > \frac{1}{3}\end{cases}\]
    \[
    	h_1(y) := [0,1], \qquad h_2(y) = \begin{cases}[0,1] & y \neq \frac{1}{2}\\ \{0\} &  y = \frac{1}{2}\end{cases}
    \]
    Using the definition of the hyperspace topology, it is straightforward to see that these maps are continuous. To check the hypothesis of the statement of causality, we must verify that the resulting two maps $A \to Y \times Z$ agree; these are just closed subsets of $[0,1]^2$. From the definition of the composition, we see that any point $(y,z)$ with $y \in g(x)$ for some $x$ and $z \in h_1(y)$ must lie in the set corresponding to $h_1$,
and similarly for $h_2$.
Hence the set corresponding to $h_1$ contains all points, while the set
corresponding to $h_2$ contains as a subset all points $(y,z)$ with $y\neq 1/2$,
as well as $(1/2,0)$.
But the closure of $\{(y,z) \mid y \neq 1/2\}$ in $[0,1]^2$ is the whole space, so that both sets are indeed equal to the whole space.

@@

Strangely enough, the only examples that we know of so far are of two kinds: first, cartesian monoidal categories (meaning that $\cC = \cC_{\det}$) with all products; and second, the finitary complete monoids of \Cref{poset_case} which are monoidal posets. For the former, the universal property of the monoidal structure implies that the causality axiom holds. For the latter, it holds trivially since any two parallel morphisms are equal.
